---
title: "Feasability Testing"
---
```{r, message=FALSE}
library(dplyr)
library(tidyr)
library(httr)
library(jsonlite)
library(ggplot2)
source("./R/functions.R")
```


## Multi-Count Query

Attmepting to get categorical subtotals for a set list of categories e.g.,

For this test I will use patient.reaction.reactionmeddrapt.exact counts within primarysource.qualification

```{r}
# get top 1000 patient.reaction.reactionmeddrapt.exact for each
topReacts = lapply(1:5, function(x){
  tempDF = getResults(list(search = paste0("primarysource.qualification:", x), count="patient.reaction.reactionmeddrapt.exact", limit=1000))
  tempDF$primarysource.qualification = x
  return(tempDF)
})
topReacts = bind_rows(topReacts) %>% left_join(loadCategoryMap("primarysource.qualification") %>% rename(primarysource.qualification = term))
str(topReacts)

```


*Top-5 Reactions Within Each Reporting Type*
```{r}
plotData = topReacts %>% group_by(term_value) %>% filter(order(-count) <= 5)
ggplot(plotData, aes(as.character(term_value), count, fill = term)) +
         geom_bar(stat = "Identity") +
  xlab("Reporter Type") +
  coord_flip()

```

*Normalized* 

```{r}
plotData = topReacts %>% group_by(term_value) %>% mutate(pct_of_total = count / sum(count)) %>% filter(order(-pct_of_total) <= 5)
ggplot(plotData, aes(as.character(term_value), pct_of_total, fill = term)) +
         geom_bar(stat = "Identity") +
  xlab("Reporter Type") +
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  coord_flip()

```



## Download Records Beyond Limit Using Skip

Find a query with a resonable number of records to download

```{r}
qList = list(
  search = 'patient.drug.medicinalproduct:"Zoladex"+AND+serious:1+AND+primarysource.qualification:5',
  limit=100)

q = createQuery(qList)
initialQuery = getResults(q, excludeMeta = FALSE)
total_results = initialQuery$meta$results$total
print(total_results)
```

Calculate number of iterations necessary and then iteratively download all records

```{r}
#Get all records if more than 100 total (TODO: should check if mutliple connections are possible and run in parallel or use source code to change limits)
if(total_results > nrow(initialQuery$results)){
  totalIterations = floor(total_results/100)
  fullList = lapply(paste0(q, "&skip=", 100*(0:totalIterations)), getResults, verbose = TRUE) # first 100 is redundant with initial query, should update in future
}

# Combine into single data frame
fullList = rbind_pages(fullList)
```

```{r}
# DF Rows should match metadata total 

nrow(fullList) == total_results
```

```{r}
head(fullList)
```

## Expaned Drug Search Using NDC Dataset

```{r}
# SET TARGET DRUG NAME
target_drug_name = "Zoladex"
```


```{r}
# prep for query
target_drug_forquery = paste0('"', gsub("\\s+", "+", target_drug_name), '"')

ndc_info = getResults(createQuery(list(search = paste0(paste0(c("brand_name:", "generic_name:"), target_drug_forquery), collapse = "+")), 
                                  rootURL = "https://api.fda.gov/drug/ndc.json?"))
label_info = getResults(createQuery(list(search = paste0(paste0(c("openfda.brand_name:", "openfda.generic_name:"), target_drug_forquery), collapse = "+")), 
                                    rootURL = "https://api.fda.gov/drug/label.json?"))

print(label_info)
print(ndc_info)

```

```{r}
#TODO: need to test how this works it there are multiple generic names not sure what the response will be shaped like
generic_names = unlist(ndc_info$generic_name)
generic_names_forquery = gsub("\\s+", "+",generic_names)
```

```{r}
qList = list(
  search =  paste0(c("openfda.brand_name:", "openfda.generic_name:", "patient.drug.medicinalproduct:"), 
                   paste0("(",paste0( c(target_drug_forquery, generic_names_forquery), collapse = "+"), ")"), collapse = "+"),
  "count" = "patient.reaction.reactionmeddrapt.exact", 
  "limit" = 10)

countPlot(getResults(qList), "Adverse Event Type", paste("Top-10 AE for Reports with", paste(c(target_drug_name, generic_names), collapse = " or ")))
```

```{r}
qList = list(
  search =  paste0(c("openfda.brand_name:", "openfda.generic_name:", "patient.drug.medicinalproduct:"),target_drug_forquery , collapse = "+"),
  "count" = "patient.reaction.reactionmeddrapt.exact", 
  "limit" = 10)

countPlot(getResults(qList), "Adverse Event Type", paste("Top-10 Reported AE for Reports with", target_drug_name))
```

```{r}
qList$count = "patient.drug.medicinalproduct.exact"
qList$limit = 20
countPlot(getResults(qList), "Reported Name", "Report Counts by Harmonized Generic Name")
```



## TODO:Automatically Download All Datasets for A Given Quarter







